---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First of all the script checks if there is a folder "data" in a working directory
and if there is a file "activity.csv" in that directory. If there is no such a file
the script will download and unzip it, if there is no such a directory - it will
create one.
```{r}
if (!file.exists("./data"))
    dir.create("./data")
if (!file.exists("./data/activity.csv"))
{
    fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(fileURL, "./data/data.zip",method = "curl")
    unzip("./data/data.zip", exdir = "./data")
}
```
Script reads the data to the data frame "data" and converts the column "date" from numeric
to Date type.
```{r}
data <- read.csv("./data/activity.csv")
data$date <- as.Date(as.character(data$date),"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
To answer this question script aggregates data on day and then counts mean and
median number of steps.
```{r}
stepsByDay <- aggregate(steps~date,data=data,sum)
meanSteps <- mean(stepsByDay$steps)
medianSteps <- median(stepsByDay$steps)
```
```{r echo = FALSE}
stringmeanSteps <- toString(round(meanSteps,2))
```
So, meanSteps = `r stringmeanSteps` and medianSteps = `r medianSteps` steps.  
After that script makes a histogram of the total number of steps taken each day.
```{r}
library(ggplot2)
range <- range(stepsByDay$steps)
bw <- (range[2] - range[1])/30
ggplot(stepsByDay,aes(steps)) + geom_histogram(binwidth = bw) + 
    geom_vline(xintercept = meanSteps,colour = "green") + 
    geom_vline(xintercept = medianSteps, colour = "red") + 
    xlab("Number of steps") + ylab("Frequency")
```

Mean and median lines overlap.

## What is the average daily activity pattern?

To find it out the script makes time series plot of average number of steps taken.
```{r}
stepsByInterval <- aggregate(steps~interval,data=data,mean)
maxTimeInt <- which.max(stepsByInterval$steps)
maxTime <- stepsByInterval$interval[maxTimeInt]
maxSteps <- stepsByInterval$steps[maxTimeInt]
maxStepsR <- round(maxSteps)
pointLabel <- paste0("(", maxTime, ", ", maxStepsR,")")

ggplot(stepsByInterval,aes(x=interval,y=steps)) + geom_line() +
    annotate("point",x=maxTime, y=maxSteps,colour = "blue") +
    annotate("text",x=maxTime + 5 , y=maxSteps + 7 ,
             label = pointLabel, size = 3) +
    xlab("Time interval") + ylab("Number of steps")
```

The plot shows also the point with maximum number of steps. So, maximum of average
number of steps is `r maxStepsR` on the interval `r maxTime`.

## Imputing missing values

There are missing values in the data set.
```{r}
colSums(is.na(data))
```
Script counts how many percents it is:
```{r}
round(100*mean(is.na(data$steps)))
```
To fix it, the script imputes mean for that 5-minute interval instead of NA.
```{r}
dataImputed <- data
for (i in 1: nrow(dataImputed))
    if(is.na(dataImputed$steps[i]))
    {
        num <-which(stepsByInterval$interval == dataImputed$interval[i])
        dataImputed$steps[i] <- stepsByInterval$steps[num]
    }
```
Script counts mean and median again.
```{r}
stepsByDay2 <- aggregate(steps~date,data=dataImputed,sum)
range2 <- range(stepsByDay2$steps)
bw2 <- (range2[2] - range2[1])/30
meanSteps2 <- mean(stepsByDay2$steps)
medianSteps2 <- median(stepsByDay2$steps)
meanSteps2
medianSteps2
```
And makes histogram of total number of steps on imputed data.
```{r}
ggplot(stepsByDay2,aes(steps)) + geom_histogram(binwidth = bw2) + 
    geom_vline(xintercept = medianSteps2, colour = "red") + 
    xlab("Number of steps") + ylab("Frequency")
```

Median is still the same, mean have changed a little bit and now equals median.
Script compares histograms.
```{r}
stepsByDay$type <- "Original"
stepsByDay2$type <- "With imputed values"
stepsByDayBig <- rbind(stepsByDay,stepsByDay2)
ggplot(stepsByDayBig,aes(steps)) + geom_histogram(binwidth = bw) + 
    geom_vline(xintercept = meanSteps,colour = "red") + facet_grid(.~type) +
    xlab("Number of steps") + ylab("Frequency")
```

## Are there differences in activity patterns between weekdays and weekends?

Script adds new factor variable "dayType" - weekday/ weekend to discover it.
```{r}
dataWeekDays <- weekdays(dataImputed$date)
dataWeekDays<-gsub("Monday","weekday",dataWeekDays)
dataWeekDays<-gsub("Tuesday","weekday",dataWeekDays)
dataWeekDays<-gsub("Wednesday","weekday",dataWeekDays)
dataWeekDays<-gsub("Thursday","weekday",dataWeekDays)
dataWeekDays<-gsub("Friday","weekday",dataWeekDays)
dataWeekDays<-gsub("Saturday","weekend",dataWeekDays)
dataWeekDays<-gsub("Sunday","weekend",dataWeekDays)
dataWeekDays <- as.factor(dataWeekDays)
dataImputed$dayType <- dataWeekDays

stepsByIntervalCompare <- aggregate(steps~interval+dayType,data=dataImputed,mean)
ggplot(stepsByIntervalCompare,aes(x=interval,y=steps)) + geom_line() +
    facet_wrap(dayType~.,nrow=2, strip.position="top") +
    xlab("Time interval") + ylab("Number of steps")
```

So, there is now such a morning peak on weekends as on weekdays, but all weekend
day is a little bit more active.